Exploring the linear separablity of syntactic and semantic information in BERT embeddings

### 2022-06-10 paper/paper.tex
fix citation parentheses

%may remove this
To achieve our goal, we construct a linear probing system for a task and then employ Iterative Nullspace Projection (INLP from here on) \citep{inlp} to generate a new embedding devoid of information learned from the probing task. We then measure the performance of this new embedding on downstream syntactic and semantic classification tasks. The design of our probing procedure follows \citep{amnesia}, which employs INLP to investigate whether BERT uses part-of-speech (POS) information when solving language modeling (LM) tasks. INLP has been used for a variety of tasks (\citep{inlp}, \citep{amnesia}, \citep{mbertGreek}, etc.) though this is the first case we know of in which INLP has been used to investigate the linear separability of syntactic and semantic information. 
%why did amnesia have citealp?

@@

\citealp{disentangle} use paraphrase pairs and new target syntax to train a semantic encoder, syntactic encoder and decoder to learn separate representations of the semantic and syntactic information contained in BART embeddings, in order to create semantically equivalent paraphrases with the new syntactic structure. Alongside the encoders they also train an adversarial syntax discriminator to try and predict the source syntax from the semantic embeddings, thus encouraging the disentanglement of the semantic and syntactic information by training the semantic embedder to remove as much syntactic information as possible. Their results show that one can achieve some removal of syntactic information from semantic embeddings, so disentanglement of some information is possible. Though they do not achieve perfect separation of the two information sets. Other non-linear approaches to syntactic-semantic information disentanglement have been carried out in \citep{multiDis}

@@

INLP, introduced in \citep{inlp}, is a method to define a linear guarding function that masks all the linear information in a word embedding that may be used for a downstream classification task. In the original paper the authors use this method to remove gender bias from BERT embeddings of biographical descriptions and then measure how easy it is to determine an individual's gender from the guarded embedding by using various downstream classification methods. Beyond using the INLP method to guard protected attributes, the authors hypothesize several additional use cases for this procedure, including information disentanglement.

The authors of \citep{amnesia} use INLP for exactly this task. They use INLP in order to separate and guard certain linguistic information sets from BERT embeddings in order to better understand what information is being used by large language models, and not just what is encoded. The main premise behind this paper is that if a particular property is used to solve a task, then the removal of that property should negatively influence the model's ability to solve that task. Conversely, if the removal of a property has little influence on the model's ability to perform a task then we know that property is not a significant contributing factor in the model's ability to perform that task. Specifically, \citep{amnesia} seeks to quantify the importance of the information sets used for part-of-speech tagging, syntactic dependency labeling, named entity recognition and syntactic constituency boundaries on BERT's ability to perform the language modeling task.

We take a similar approach to \citep{amnesia} by separating the information sets used for CCG tagging and semantic labeling from word-level BERT embeddings. However, as we are interested in the linear separability of these two information sets, we will test how the removal of these information sets impacts the embeddings' performance on semantic labeling and CCG tagging, respectively, rather than language modeling.

@@

The INLP method first introduced in \citep{inlp}, is used to create a guarding function that masks all the linear information contained in a set of vectors, $X$, that can be used map each vector to $c \in C$, where $C$ is the set of all categories. This is accomplished by training a linear classifier, a matrix $W$, that is applied to each $x \in X$ in order to predict the correct category $c$ with the greatest possible accuracy. In other words, $Wx$ defines a distribution over the set of categories $C$ and we assign $x$ to the class $c \in C$ which is allotted the greatest probability by $Wx$. Note that the classifier's accuracy must be greater than that achieved by guessing the majority category, otherwise $x$ contains no linear information relevant for the categorization task and thus no guarding function is needed. Once $W$ is determined, for any $x \in X$ we can remove the information that $W$ uses to predict $c$ by projecting $x$ onto the null-space of $W$, $N(W) = \{x | Wx=0\}$. Call this projection function $P_1$ and let $\hat{x} = P_1(x)$. This removes all of the linear information in $x$ that $W$ used to predict the category $c$. 
%Need to cite INLP paper

>>>>

%may remove this
To achieve our goal, we construct a linear probing system for a task and then employ Iterative Nullspace Projection (INLP from here on) \citep{inlp} to generate a new embedding devoid of information learned from the probing task. We then measure the performance of this new embedding on downstream syntactic and semantic classification tasks. The design of our probing procedure follows \citealp{amnesia}, which employs INLP to investigate whether BERT uses part-of-speech (POS) information when solving language modeling (LM) tasks. INLP has been used for a variety of tasks (\citealp{inlp}; \citealp{amnesia}; \citealp{mbertGreek}, etc.) though this is the first case we know of in which INLP has been used to investigate the linear separability of syntactic and semantic information. 
%why did amnesia have citealp?

@@

\citealp{disentangle} use paraphrase pairs and new target syntax to train a semantic encoder, syntactic encoder and decoder to learn separate representations of the semantic and syntactic information contained in BART embeddings, in order to create semantically equivalent paraphrases with the new syntactic structure. Alongside the encoders they also train an adversarial syntax discriminator to try and predict the source syntax from the semantic embeddings, thus encouraging the disentanglement of the semantic and syntactic information by training the semantic embedder to remove as much syntactic information as possible. Their results show that one can achieve some removal of syntactic information from semantic embeddings, so disentanglement of some information is possible. Though they do not achieve perfect separation of the two information sets. Other non-linear approaches to syntactic-semantic information disentanglement have been carried out in \citealp{multiDis}

@@

INLP, introduced in \citealp{inlp}, is a method to define a linear guarding function that masks all the linear information in a word embedding that may be used for a downstream classification task. In the original paper the authors use this method to remove gender bias from BERT embeddings of biographical descriptions and then measure how easy it is to determine an individual's gender from the guarded embedding by using various downstream classification methods. Beyond using the INLP method to guard protected attributes, the authors hypothesize several additional use cases for this procedure, including information disentanglement.

The authors of \citealp{amnesia} use INLP for exactly this task. They use INLP in order to separate and guard certain linguistic information sets from BERT embeddings in order to better understand what information is being used by large language models, and not just what is encoded. The main premise behind this paper is that if a particular property is used to solve a task, then the removal of that property should negatively influence the model's ability to solve that task. Conversely, if the removal of a property has little influence on the model's ability to perform a task then we know that property is not a significant contributing factor in the model's ability to perform that task. Specifically, \citealp{amnesia} seeks to quantify the importance of the information sets used for part-of-speech tagging, syntactic dependency labeling, named entity recognition and syntactic constituency boundaries on BERT's ability to perform the language modeling task.

We take a similar approach to \citealp{amnesia} by separating the information sets used for CCG tagging and semantic labeling from word-level BERT embeddings. However, as we are interested in the linear separability of these two information sets, we will test how the removal of these information sets impacts the embeddings' performance on semantic labeling and CCG tagging, respectively, rather than language modeling.

@@

The INLP method first introduced in \citealp{inlp}, is used to create a guarding function that masks all the linear information contained in a set of vectors, $X$, that can be used map each vector to $c \in C$, where $C$ is the set of all categories. This is accomplished by training a linear classifier, a matrix $W$, that is applied to each $x \in X$ in order to predict the correct category $c$ with the greatest possible accuracy. In other words, $Wx$ defines a distribution over the set of categories $C$ and we assign $x$ to the class $c \in C$ which is allotted the greatest probability by $Wx$. Note that the classifier's accuracy must be greater than that achieved by guessing the majority category, otherwise $x$ contains no linear information relevant for the categorization task and thus no guarding function is needed. Once $W$ is determined, for any $x \in X$ we can remove the information that $W$ uses to predict $c$ by projecting $x$ onto the null-space of $W$, $N(W) = \{x | Wx=0\}$. Call this projection function $P_1$ and let $\hat{x} = P_1(x)$. This removes all of the linear information in $x$ that $W$ used to predict the category $c$. 
%Need to cite INLP paper
